A general and efficient representation of Ancestral Recombination Graphs

### 2023-08-30 paper.tex
Update

\section*{Levels of simplification}\label{ARG_simplification}
ARG simplification is a powerful tool, and can do much more
besides removing unreachable nodes and edges in a prospective ARG
and resolving ancestral material in an eARG.
In general, we can think of
ARG simplification as the process
of removing nodes and re-writing edges in a gARG to remove
various types of redundancy.
Many of the types of redundancy revolve around the presence
of ``unary'' nodes in the local trees
(a tree node that has exactly one child).

@@

Fig.~\ref{fig-simplification} show a series of successive steps of
simplification, starting with a complete [TODO be more concrete about
what is stored here; ``complete'' isn't very helpful] gARG simulated under a
backwards-in-time Wright-Fisher
model, shown both as a graph on the left, and the corresponding sequence of
local trees on the right.
In Fig.~\ref{fig-simplification}A the local trees (right)
contain many unary nodes, and as we simplify more and more
the local trees contain fewer
unary nodes (Fig.~\ref{fig-simplification}B,C) until
we reach Fig.~\ref{fig-simplification}D in which the local trees
have no unary nodes.
Unary nodes in a local tree arise from
genetic inheritance
passing through an ARG node in which no coalescence
occurs in the corresponding genome interval.
The distinction between the ``common ancestry'' of two or more genomes
in an ancestral genome and the ``coalescence'' which may or may
not occur in the local trees is
important~\citep{hudson1983testing,kelleher2016efficient}.
Consider \noderef{e} in in Fig.~\ref{fig-simplification}A,
for example. We can see from the graph that it is a common
ancestor of  samples \noderef{a} and \noderef{b}, but
it does not correspond to any coalescence in the
local trees to the left of position $44$, and is a unary
node in these three trees. Unary nodes in the local trees
arise from a variety of processes, and would not usually
be considered significant. We are usually interested in tree
nodes that have at least two children, because these
are in principle detectable from mutational differences
arising in the respective subtrees.

@@

\section*{Precision of recombination information}
% The ARGs in Fig.~\ref{fig-simplification} represent different
% levels of detail about the same ancestral history. They represent
% the same set of recombination and common ancestor events,
As illustated in Fig.~\ref{fig-simplification}, successive levels
of ARG simplification reduce the amount of information about the
history of the sample that is stored. Some of the information lost,
for example, ``diamond'' removal (Fig.~\ref{fig-simplification}A),
seems like a reasonable tradeoff for a simpler structure.
The consequences of other simplifications, however, are
more subtle and relate directly to what can be known about
recombination events and the levels of precision that
we should seek to infer about them.

@@

@@

The point of these examples is to emphasise that
information about recombination in an ARG is not ``all
or nothing'': there is a gradation of detail
about recombination, in terms of genomic location, timing,
and the ancestral lineages involved.
While there has been a recent tendancy to emphasise the
importance of ``full ARGs''
\citep[e.g.][]{deng2021distribution,brandt2021evaluation,
rasmussen2022espalier}
which contain complete information about all recombination
events, it is reasonable to question the value of
such precise information and whether it is knowable.
[SOME MORE STUFF]

@@

\begin{figure} \begin{center}
\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
\caption{\label{fig-inferred-args} ARGs inferred by
(A) \kwarg~\citep{ignatieva2021kwarg},
(B) \argweaver~\citep{rasmussen2014genome,hubisz2020inference},
(C) \tsinfer~\citep{kelleher2019inferring},
and (D) \relate~\citep{speidel2019method}
for 11 samples of a 2.7kb region of the \textit{Drosophila melanogaster} ADH
locus~\citep{kreitman1983nucleotide}. The \argweaver\ example
is a sample from the inferred posterior distribution. In each ARG,
colours are assigned to edges based on the time of the edge's child node:
lighter colours are assigned to edges with older child
nodes, darker colours are assigned to edges with younger child nodes
Note that the position of points on the
Y axis is chosen for graphical clarity by the graphviz \texttt{dot}
algorithm, rather than as a direct measure of time.
Line width corresponds to the genomic span of the edge. Node colour
refers to the number of parents of the node: red is zero, brown is one,
green to blue are two to N parents. The brightness
of the node is determined by the genomic span over which
the node has two or more children divided by the span where it has any
children. The genomic intervals associated with edges are visualised below
each ARG, where each row is an edge
and the x axis reflects genomic position. The width of edges
thus corresponds to their genomic span. Edges
are ordered by the age of their child node, from oldest to youngest
moving down the page, with colours matching the edge colours in the ARG.
Vertical lines indicate breakpoints between local trees.
}
\end{figure}

To illustrate these points and to emphasise the heterogeneity
of the ARGs inferred by current methods, we inferred ARGs for the
the classical \citet{kreitman1983nucleotide} benchmark dataset
using four recent inference methods and show the results in
Fig~\ref{fig-inferred-args}.
While there is some consistency among the inferred ARGs (such as the samples
FR-F, WA-F and Af-F being grouped into the same clade), there
are also considerable differences. For this dataset, the minimum number
of recombination breakpoints required in the absence of recurrent mutation
is 7~\citep{song2003parsimonious}, and the number of inferred breakpoints
under the input mutation and recombination rates
varies from 2 (\relate) to 17 (\argweaver).
The ARGs inferred by the different methods also have major differences
in how recombination information is represented. Essentially,
\kwarg\ and \argweaver\ infer a specific causal recombination \emph{event}
for each breakpoint, and those events are fully
described in the output graph. On the other hand, \tsinfer\ and
\relate\ do not infer causal events but instead focus on
representing the inferred \emph{effect}: breakpoints between the
trees. It is important to note that this does not imply that
the methods must then necessarily infer only a sequence of
\emph{unrelated} local trees (i.e., with either unlabelled internal
nodes or differently labelled nodes in each tree; see the
next section).
The ARGs inferred by \relate\ and \tsinfer\ both contain substantial
information about node sharing across trees, and hence capturing
the important correlation structures.

% \relate\ by first inferring local topologies and then identifies edges
% that persist from one tree, while
% local trees only emerge from \tsinfer\ as a consequence of the copying
% patterns inferred among ancestral haplotypes.

>>>>

\section*{Levels of simplification}\label{ARG_simplification}
ARG simplification is a powerful tool, and can do much more
besides removing unreachable nodes and edges in a prospective ARG
and resolving ancestral material in an eARG.
In general, we can think of
ARG simplification as the process
of removing nodes and re-writing edges (and their inheritance annotations)
to remove various types of redundancy.
The redundancy that we are interested in 
revolves around the presence
of ``unary'' nodes in the local trees
(a tree node that has exactly one child), which arise in several different ways.
Unary nodes are not a standard feature of evolutionary trees;
we are usually interested
nodes that have at least two children, because these
are in principle detectable from mutational differences
arising in the respective subtrees.
A tree branch may correspond to many generations 
of individuals (or indeed cell divisions) and there is usually
very little information about these intermediates, or utility
in modelling them. 
% FIXME the bridge between the next few paragraphs is a bit weird, in 
% that we've just been talking about how interesting unary nodes are
% but then go on about how to remove them. Just needs a bit of smoothing
% out I think.
ARGs, however, are different, and there 
can be significant information about evolutionary processes
provided by unary nodes in the local trees.

@@

Fig.~\ref{fig-simplification} show a series of successive steps of
simplification, starting with a complete [TODO be more concrete about
what is stored here; ``complete'' isn't very helpful] gARG simulated under a
backwards-in-time Wright-Fisher
model, shown both as a graph on the left, and the corresponding sequence of
local trees on the right.
In Fig.~\ref{fig-simplification}A the local trees (right)
contain many unary nodes, and as we successively simplify
the local trees contain fewer
unary nodes (Fig.~\ref{fig-simplification}B,C) until
we reach Fig.~\ref{fig-simplification}D, where the local trees
have no unary nodes.
A node is unary in a local tree covering the genome interval $I$
if genetic inheritance passes through that ARG node,
but no coalescence occurs in the interval $I$.
The distinction between the ``common ancestry'' of two or more genomes
in an ancestral genome and the ``coalescence'' which may or may
not occur in the local trees is
important~\citep{hudson1983testing,kelleher2016efficient}.
Consider \noderef{e} in in Fig.~\ref{fig-simplification}A,
for example. We can see from the graph that it is a common
ancestor of  samples \noderef{a} and \noderef{b}, but
it does not correspond to any coalescence in the
local trees to the left of position $44$, and is a therefore a 
unary node in these three trees.

@@

\section*{Precision of recombination information}
% The ARGs in Fig.~\ref{fig-simplification} represent different
% levels of detail about the same ancestral history. They represent
% the same set of recombination and common ancestor events,
As illustated in Fig.~\ref{fig-simplification}, successive levels
of ARG simplification reduce the amount of information about the
history of the sample that is stored. Some of the information lost,
e.g.\ ``diamond'' removal (Fig.~\ref{fig-simplification}A),
seems like a reasonable tradeoff for a simpler structure.
The consequences of other simplifications, however, are
more subtle and relate directly to what can be known about
recombination events and the levels of precision that
we should seek to infer about them.

@@

% or nothing'': there is a gradation of detail
% about recombination, in terms of genomic location, timing,
% and the ancestral lineages involved.
% that vary levels of precision about reccomb

@@

% The point of these examples is to emphasise that
% information about recombination in an ARG is not ``all
% or nothing'': there is a gradation of detail
% about recombination, in terms of genomic location, timing,
% and the ancestral lineages involved.
% While there has been a recent tendancy to emphasise the
% importance of ``full ARGs''
% \citep[e.g.][]{deng2021distribution,brandt2021evaluation,
% rasmussen2022espalier}
% which contain complete information about all recombination
% events, it is reasonable to question the value of
% such precise information and whether it is knowable.
% [SOME MORE STUFF]

@@

\begin{figure} \begin{center}
\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
\caption{\label{fig-inferred-args} ARGs inferred by
(A) \kwarg, (B) \argweaver, (C) \tsinfer,
and (D) \relate\
for 11 samples of a 2.7kb region of the \textit{Drosophila melanogaster} ADH
locus~\citep{kreitman1983nucleotide}. 
The \argweaver\ example
is a sample from the inferred posterior distribution. In each ARG,
colours are assigned to edges based on the time of the edge's child node
(lighter: older; darker: younger).
Position on the X and Y axes for the graph visualization is arbitrary.
% Note that the position of points on the
% Y axis is chosen for graphical clarity by the graphviz \texttt{dot}
% algorithm, rather than as a direct measure of time.
Line width corresponds to the genomic span of the edge. Node colour
refers to the number of parents of the node.
The fill colour intensity of a node is determined by the genomic span over which
it has two or more children divided by the span where it has any
children (the ``coalescent span''). 
The genomic intervals associated with edges are shown below
each ARG, where each row is an edge
and the X axis reflects genomic position. 
% Edges
% are ordered by the age of their child node, from oldest to youngest
% moving down the page, with colours matching the edge colours in the ARG.
Vertical lines indicate breakpoints between local trees.
}
\end{figure}

The example of successive simplification of a simulated ARG
in Fig.~\ref{fig-simplification} illustrates the point
that is is possible to \emph{store} varying levels of precision
about recombination in an ARG. 
It is also possible, of course, to \emph{infer} varying levels of precision about 
recombination.
To illustrate this, and to demonstrate the structural 
heterogeneity of the ARGs inferred  by current methods,
Fig.~\ref{fig-inferred-args} shows the results ARG inference
% To illustrate these points and to emphasise the heterogeneity
% of the ARGs inferred by current methods, we inferred ARGs for the
for the classical \citet{kreitman1983nucleotide} benchmark dataset
using  \kwarg~\citep{ignatieva2021kwarg},
\argweaver~\citep{rasmussen2014genome,hubisz2020inference},
\tsinfer~\citep{kelleher2019inferring},
and \relate~\citep{speidel2019method}.
While there is some consistency among the inferred ARGs (such as the samples
FR-F, WA-F and Af-F being grouped into the same clade), there
are also considerable differences. For this dataset, the minimum number
of recombination breakpoints required in the absence of recurrent mutation
is 7~\citep{song2003parsimonious}, and the number of inferred breakpoints
under the input mutation and recombination rates
varies from 2 (\relate) to 17 (\argweaver).
% TODO this needs a bit of rewriting with the rest of the paper in mind
The ARGs inferred by the different methods also have major differences
in how recombination information is represented. 
Essentially,
\kwarg\ and \argweaver\ infer a specific causal recombination \emph{event}
for each breakpoint, and those events are fully
described in the output graph. On the other hand, \tsinfer\ and
\relate\ do not infer causal events but instead focus on
representing the inferred \emph{effect}: breakpoints between the
trees. It is important to note that this does not imply that
the methods must then necessarily infer only a sequence of
\emph{unrelated} local trees (i.e., with either unlabelled internal
nodes or differently labelled nodes in each tree; see the
next section).
The ARGs inferred by \relate\ and \tsinfer\ both contain substantial
information about node sharing across trees, and hence lie somewhere 
in between a sequence of unrelated trees on one extreme and a 
completely precise event ARG on the other, in terms of the precision
inferred about recombination.

% \relate\ "postorder" first inferring local topologies and then identifies edges
% that persist from one tree, while
% local trees only emerge from \tsinfer\ as a consequence of the copying
% patterns inferred among ancestral haplotypes.
